Formalization of Network Flow Algorithms

### 2017-03-03 paper/root.tex
Reza's comments, more precise citation of Cherkassky-Goldberg reference implementation

Using stepwise refinement techniques, we instantiate the generic Ford-Fulkerson method to Edmonds-Karp algorithm, 
and the generic push-relabel algorithm of Goldberg and Tarjan to both the relabel-to-front and the FIFO push-relabel algorithm.
Further refinement then yields verified efficient implementations of the algorithms, which compare well to unverified reference implementations.

@@

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum bipartite matching, edge disjoint paths, circulation demand, as well as various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. 
The basic idea is to repeatedly search for \emph{augmenting paths} along which more flow can be moved from the source to the sink.
An important instance is the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time ($O(VE^2)$) for the general case of networks with real-valued capacities.
Dinitz~\cite{Di06} generalized the idea of augmenting paths to blocking flows, obtaining an $O(V^2E)$ algorithm. 
Karzanov~\cite{Ka74} was the first to propose an $O(V^3)$ algorithm based on the idea of \emph{preflows}. Based on Karzanov's ideas, Goldberg and Tarjan proposed
the generic push-relabel algorithm~\cite{GoTa88}.
Implementations of the push-relabel algorithm are among the most efficient maximum flow algorithms. While the generic algorithm has a time complexity of $O(V^2E)$,
specific variants of the algorithms achieve even lower complexities down to $O(V^2\sqrt{E})$.

@@

This paper is an extended version of our conference paper~\cite{LaSe16}, which only covered the Edmonds-Karp algorithm.
While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial time maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation. Moreover, this paper is a case study on elegantly formalizing algorithms.

In the rest of this paper, we give a short informal description the Ford-Fulkerson method (\S\ref{sec:background})
and present our formalization of it (\S\ref{sec:abs-formalization}). After introducing the Isabelle Refinement Framework~\cite{LaTu12,La12} (\S\ref{sec:refinement}), we report on the instantiation of the Ford-Fulkerson method to the Edmonds-Karp algorithm, and on its complexity proof (\S\ref{sec:edka}).
Next, we describe the correctness and complexity proof of the generic push-relabel algorithm (\S\ref{sec:gen-prpu}), and its instantiations 
to the relabel-to-front and FIFO push-relabel algorithms (\S\ref{sec:prpu-inst}).
Finally, we report on the refinement of all three algorithms to efficient implementations (\S\ref{sec:executable}), and their comparison to unverified reference 
implementations (\S\ref{sec:benchmark}).

The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/edmonds_karp/}. 
The part on Edmonds-Karp algorithm is also available in the Archive of Formal Proofs~\cite{LaSe16_afp}.

@@

\section{The Ford-Fulkerson Method}\label{sec:background}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}

In this section, we give a short introduction to the Ford-Fulkerson method, closely following the presentation by Cormen et al.~\cite{CLRS09}.

@@

A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all nodes except $s$ and $t$, the sum of incoming flows equals the 
sum of outgoing flows.
% 
% \emph{Non-generation constraint}: For all nodes except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
% We define the \emph{excess} flow $x_f(u)$ of a node $u$ to be the difference of the incoming and outgoing flow of $u$.
% Intuitively, a \emph{preflow} must not send more flow over an edge than its capacity allows, and only the source node may generate flow. 
% The excess flow is the amount of flow that is stuck at a node, \ie which flows in but does not flow out.
% A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all nodes except $s$ and $t$, the excess flow is zero.
% Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

@@

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges in the same and opposite direction as the network edges.
% , which has edges both in parallel and reverse to the edges of the network. 
Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

In each iteration, the Ford-Fulkerson method tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph.
It then pushes as much flow as possible along this path to increase the value of the current flow. 
Formally, for an augmenting path $p$, one first defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
An augmenting path then yields a residual flow $f_p$, which is the flow that can be passed along this path:
\[ f_p (u, v) = 
  \begin{cases}
  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
\newcommand{\augment}{{\mathbin\uparrow}}%
Finally, to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
\[ (f \augment f') (u, v) = 
  \begin{cases}
  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
Note that, for any edge in the network, the augmenting flow in the same direction is added to the flow, while the augmenting flow in the opposite direction is subtracted. 
This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.

@@

The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph. There are several possible implementations with different execution times. The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}. When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real-valued capacities. Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges if the intended meaning is clear from the context.
A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.

\section{Formalizing the Ford-Fulkerson Method}\label{sec:abs-formalization}
%  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
%    Reader should think: Yeah, they did cool stuff!
%
%  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
%      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]
%    
%  Proof of MinCut-MaxFlow: Follows the textbook proof.
%    Uses Isar to even look like a textbook proof, being comprehensible even without running
%      Isabelle/HOL.
%      
%    For example, \isai{augment_flow_presv_cap}.
%      Present formal proof text. Perhaps oppose it to textbook proof?.
%    
%  Abstract Algo looks like pseudocode presented in texbooks.
%    Oppose our algo to textbook pseudocode.
%

@@

In this section, we provide a brief overview of our formalization of the Ford-Fulkerson method. 
In order to develop theory in the context of a fixed graph or network, we use Isabelle's concept of \emph{locales}~\cite{Ballarin:2006:MKM}, which allows
us to define named contexts that fix some parameters and assumptions.
For example, the graph theory is developed in the locale \isai{Graph}, which 
fixes the edge labeling function \isai{c}, and defines the set of edges and nodes based on $c$:
\begin{lstlisting}
  locale Graph = fixes c :: "edge => capacity" begin
    definition "E == {(u, v). c (u, v) ~= 0}"
    definition "V == {u. \<exists>v. (u, v) \<in> E \<or> (v, u) \<in> E}"
    [...]
\end{lstlisting}
Moreover, we define basic concepts like (simple, shortest) paths, and provide lemmas to reason about them.

@@

For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that resemble standard mathematical textbook proofs, and are accessible, to a certain extent, even for those not familiar with Isabelle/HOL. 
We use Isar to present our proof of the Ford-Fulkerson method such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.

@@

\subsection{Presentation of Algorithms}
In textbooks, it is common to present algorithms in pseudocode, which captures the essential ideas, but leaves open implementation details. As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}. For example, we define the Ford-Fulkerson method as follows:
\Snippet{ford_fulkerson_algo}

@@

Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Isabelle Refinement Framework, it is straightforward to prove (partial) correctness of
the Ford-Fulkerson method, which is stated in Isabelle/HOL by the following theorem:
\Snippet{ford_fulkerson_correct}

\section{Refinement in Isabelle/HOL}\label{sec:refinement}
After having stated and proved correct an algorithm on the abstract level, the next step is to provide an (efficient) implementation. 
In our case, we first specialize the Ford-Fulkerson method to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.

@@

\section{The Edmonds-Karp Algorithm}\label{sec:edka}
  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm is straightforward, as finding a shortest augmenting path is a refinement of finding any augmenting path.

@@

  The push-relabel algorithm maintains a preflow $f$ and a height labeling $l: V \to \mathbb {N}$ of the nodes, 
  with the invariant that for every edge $(u, v)$ of the residual graph we have that $l(u) \le l(v) + 1$. 
  Moreover, the height of the source is fixed to $\left\vert{V}\right\vert$, and the height of the sink is fixed to $0$.
  By generalizing the labeling invariant to paths, we get that for every path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
  This implies that there cannot be a path from $s$ to $t$: the length of any (simple) path is less than $|V|$, however, a valid labeling 
  implies a length greater than or equal to $|V|$. Upon termination of the algorithm, there will be no active nodes. 
  Thus, the preflow is actually a flow, and as we have no $s$--$t$ paths, the Ford-Fulkerson theorem implies that it is maximal.

@@

The simple structure of the generic algorithm allowed us to conveniently define it as labeled transition system, such that the above informal proof could be formalized
by inductive arguments over the runs of the algorithm.
Moreover, the powerful Isabelle/HOL standard library, in particular the support for summations
over finite sets, was very helpful to keep the formalization effort manageable. 
Anyway, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few weeks to complete.
This includes the time that we spent on an initial termination proof that used simpler arguments, but did not derive an exact complexity bound.

@@

\subsection{FIFO Push-Relabel Algorithm}
While the relabel-to-front algorithm has a theoretical worst case complexity of $O(V^3)$, it seems to be rather inefficient in practice.\footnote{We have not found any implementation based on relabel-to-front, and our own experiments indicate a rather poor performance.}
An alternative implementation is the FIFO push-relabel algorithm, which maintains a first-in-first-out queue of the active nodes. 
In each iteration, it dequeues a node from the FIFO, and tries to discharge it by only using push operations.
If the node is still active, it is relabeled and enqueued.

The FIFO push-relabel algorithm has several advantages: It's correctness is straightforward to prove, 
as the only required additional invariant is that the queue contains exactly the active nodes. 
Moreover, it performs well in practice, and we have found several reference implementations of it~\cite{ChGo97,ICPC-notebook}.

@@

  \subsection{Algorithms}
  The most complex problem for which we specified no algorithm yet is finding a shortest augmenting path.
  We use breadth first search (BFS) for this. We adapted a BFS formalization from the examples collection of the Isabelle Refinement Framework to
  our needs and added an efficient imperative implementation. The resulting algorithm 
  is independent of flow networks and can be reused for other applications.

@@

  We implement capacities by (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers. Thus, we could chose any executable number representation here.}. Note that an implementation by fixed precision numbers would also be possible,
  but requires additional checks on the network to ensure that no overflows can happen.\footnote{Actually, the unverified reference 
  implementations of the push-relabel algorithms~\cite{ChGo97,ICPC-notebook} that we use in our benchmarks (\cf Section~\ref{sec:benchmark}) lack such checks, and silently report erroneous maximum flow values in case of overflow.}
  % TODO: Can we trick the Java reference implementation on integer overflow?
  Nodes are represented by natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
  are realized as arrays of size $N\times N$ with row-major indexing, such that the successors of a node are close together in memory.
  The adjacency map of the network is implemented by an array of lists of nodes. 
  An augmenting path is represented by a list of edges, \ie a list of pairs of nodes. 
  The height frequency map and the maps from nodes to height labels and excess values are implemented by arrays.
  The queue of the FIFO push-relabel algorithm and the list of the relabel-to-front algorithm are implemented by two stacks, allowing for amortized constant time operations.

@@

  For the push-relabel algorithm, we use the C++ implementation from the Stanford ACM-ICPC notebook~\cite{ICPC-notebook}, which served as 
  a template for our verified FIFO push-relabel algorithm. Moreover, we use a highly optimized C implementation of a FIFO push-relabel 
  algorithm with various additional heuristics by Cherkassky et al.~\cite{ChGo97}. These were compiled using gcc 5.4 at optimization level 4.
  The data set consists of networks from the DIMACS implementation challenge~\cite{JGC93}: 
  The ``Goldberg bad case networks'' from the ``Washington'' generator, which are tailored to be hard for push-relabel algorithms, 
  the ``fully dense acyclic networks'' of Waissi, and the RMFGEN networks of Goldfarb and Grigoriadis.

@@

@@

  For the push-relabel algorithms, our verified algorithm is, depending on the test case, between 5 and 10 times slower than the unverified C++ implementation.
  The highly optimized C implementation is up to two orders of magnitude faster. 
  On the ``goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms, 
  even Edmonds-Karp algorithm is competitive to the simple FIFO push-relabel algorithms, while the highly optimized one shows no weaknesses on these networks.

>>>>

Using stepwise refinement techniques, we instantiate the generic Ford-Fulkerson algorithm to Edmonds-Karp algorithm, 
and the generic push-relabel algorithm of Goldberg and Tarjan to both the relabel-to-front and the FIFO push-relabel algorithm.
Further refinement then yields verified efficient implementations of the algorithms, which compare well to unverified reference implementations.

@@

\section{Introduction}
Computing the maximum flow in a network is an important problem in graph theory.
Many other problems, like maximum bipartite matching, edge disjoint paths, circulation demand, as well as various scheduling and resource allocating problems can be reduced to it.
The generic Ford-Fulkerson algorithm~\cite{FF56} describes a class of algorithms to solve the maximum flow problem. 
The basic idea is to repeatedly search for \emph{augmenting paths} along which more flow can be moved from the source to the sink.
An important instance is the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time ($O(VE^2)$) for the general case of networks with real-valued capacities.
Dinitz~\cite{Di06} generalized the idea of augmenting paths to blocking flows, obtaining an $O(V^2E)$ algorithm. 
Karzanov~\cite{Ka74} was the first to propose an $O(V^3)$ algorithm based on the idea of \emph{preflows}. Based on Karzanov's ideas, Goldberg and Tarjan proposed
the generic push-relabel algorithm~\cite{GoTa88}.
Implementations of the push-relabel algorithm are among the most efficient maximum flow algorithms. While the generic algorithm has a time complexity of $O(V^2E)$,
specific variants of the algorithms achieve even lower complexities down to $O(V^2\sqrt{E})$.

@@

This paper is an extended version of our conference paper~\cite{LaSe16}, which only covered the Edmonds-Karp algorithm.
While there exists another formalization of the Ford-Fulkerson algorithm in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial time maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation. Moreover, this paper is a case study on elegantly formalizing algorithms.

In the rest of this paper, we give a short informal description of the Ford-Fulkerson algorithm (\S\ref{sec:background})
and present our formalization of it (\S\ref{sec:abs-formalization}). After introducing the Isabelle Refinement Framework~\cite{LaTu12,La12} (\S\ref{sec:refinement}), we report on the instantiation of the Ford-Fulkerson algorithm to the Edmonds-Karp algorithm, and on its complexity proof (\S\ref{sec:edka}).
Next, we describe the correctness and complexity proof of the generic push-relabel algorithm (\S\ref{sec:gen-prpu}), and its instantiations 
to the relabel-to-front and FIFO push-relabel algorithms (\S\ref{sec:prpu-inst}).
Finally, we report on the refinement of all three algorithms to efficient implementations (\S\ref{sec:executable}), and their comparison to unverified reference 
implementations (\S\ref{sec:benchmark}).

The source code of our formalization is available at \url{http://www21.in.tum.de/~lammich/max_flow/}. 
The part on Edmonds-Karp algorithm is also available in the Archive of Formal Proofs~\cite{LaSe16_afp}.

@@

\section{The Ford-Fulkerson Algorithm}\label{sec:background}
% \begin{comment}
%   GOAL: 
%     1) Remind the reader of FoFu. The educated reader should have the feeling:
%       Yes, now I recall FoFu, and understand (again) what it does.
%       
%     2) Set the field for further, more detailed descriptions in next section.
%   
% 
%   Short description of Network: Finite graph over nodes V, 
%     edge (u,v) annotated with capacity c(u,v).
%     Assuming distinct nodes s and t. 
%     Note on additional assumptions: 
%         No parallel edges, s only outgoing, t only incoming, all nodes on path from s to t. 
%         Can transform any Network to match these assumptions, preserving the flow cite{???}.
%         
%   s-t Flow: Annotation of edges with values, such that: 
%     1) Values smaller than capacities.
%     2) Kirchhoff law: sum of incoming flows = sum of outgoing flows, for all nodes except s t
%   
%   Value of flow: Incoming flow - outgoing flow of s
%   
%   Max-Flow problem. 
%     Rpt. why important?
%   
%   Cuts, intuition of min-cut >= max-flow
%   
%   Theorem: Min-Cut = Max-Flow
%   Proven via augmenting flow in residual graph. 
%     Present 1,2,3 of min-cut max-flow equivalences
%   
%   This yields an algorithm to compute max-flows, if we can compute flows in residual graph.
%     --> Simple way: Augmenting flow via augmenting path. ==> FoFu-scheme. 
%       Termination? In General: Only for integer capacities.
%       Edmonds-Karp: Shortest augmenting path. Always terminates! Complexity: O(|E||V|) outer loop iterations, BFS requires O|E|.
%   \end{comment}

In this section, we give a short introduction to the Ford-Fulkerson algorithm, closely following the presentation by Cormen et al.~\cite{CLRS09}.

@@

A \emph{flow} $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all nodes except $s$ and $t$, the sum of incoming flows equals the 
sum of outgoing flows.
% 
% \emph{Non-generation constraint}: For all nodes except $s$ and $t$, the sum of flows over all incoming edges is not smaller than the sum of flows over all outgoing edges.
% We define the \emph{excess} flow $x_f(u)$ of a node $u$ to be the difference of the incoming and outgoing flow of $u$.
% Intuitively, a \emph{preflow} must not send more flow over an edge than its capacity allows, and only the source node may generate flow. 
% The excess flow is the amount of flow that is stuck at a node, \ie which flows in but does not flow out.
% A \emph{flow} is a \emph{preflow} that satisfies the \emph{Conservation constraint}: For all nodes except $s$ and $t$, the excess flow is zero.
% Intuitively, only the source may generate flow, and only the sink may consume flow, and for all other nodes, the incoming flow must be equal to the outgoing flow.
The value of a flow $f$ is denoted by $|f|$, and defined to be the sum over the outgoing flows of $s$ minus the sum over the incoming flows of $s$.
Given a network $G$, the maximum flow problem is to find a flow with the maximum value among all flows of the network.

@@

The Ford-Fulkerson algorithm is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson algorithm defines the residual graph, which has edges in the same and opposite direction as the network edges.
% , which has edges both in parallel and reverse to the edges of the network. 
Each edge is labeled by the amount of flow that can be effectively passed along this edge, by either increasing or decreasing the flow on a network edge.
Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

In each iteration, the algorithm tries to find an \emph{augmenting path}, \ie a simple path from $s$ to $t$ in the residual graph.
It then pushes as much flow as possible along this path to increase the value of the current flow. 
Formally, for an augmenting path $p$, one first defines the \emph{residual capacity} $c_p$ as the minimum value over all edges of $p$:
\[c_f(p) = \min \{c_f(u, v): \text{$(u, v)$ is on  $p$}\}\]
An augmenting path then yields a residual flow $f_p$, which is the flow that can be passed along this path:
\[ f_p (u, v) = 
  \begin{cases}
  c_f(p) \hfill & \text{ if $(u, v)$ is on $p$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
\newcommand{\augment}{{\mathbin\uparrow}}%
Finally, to actually push the flow induced by an augmenting path, we define the \emph{augment} function $f\augment f'$, which augments a flow $f$ in the network 
by any \emph{augmenting flow} $f'$, \ie any flow in the residual graph:
\[ (f \augment f') (u, v) = 
  \begin{cases}
  f (u, v) + f'(u, v) - f'(v, u) \hfill & \text{ if $(u, v) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]
Note that, for any edge in the network, the augmenting flow in the same direction is added to the flow, while the augmenting flow in the opposite direction is subtracted. 
This matches the intuition of passing flow in the indicated direction, by either increasing or decreasing the flow of an edge in the network.

@@

The Ford-Fulkerson algorithm is generic in that it does not specify how to find an augmenting path in the residual graph. There are several possible implementations with different execution times. The generic algorithm is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}. When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real-valued capacities. Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges if the intended meaning is clear from the context.
A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.

\section{Formalizing the Ford-Fulkerson Algorithm}\label{sec:abs-formalization}
%  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
%    Reader should think: Yeah, they did cool stuff!
%
%  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
%      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]
%    
%  Proof of MinCut-MaxFlow: Follows the textbook proof.
%    Uses Isar to even look like a textbook proof, being comprehensible even without running
%      Isabelle/HOL.
%      
%    For example, \isai{augment_flow_presv_cap}.
%      Present formal proof text. Perhaps oppose it to textbook proof?.
%    
%  Abstract Algo looks like pseudocode presented in texbooks.
%    Oppose our algo to textbook pseudocode.
%

@@

In this section, we provide a brief overview of our formalization of the Ford-Fulkerson algorithm. 
In order to develop theory in the context of a fixed graph or network, we use Isabelle's concept of \emph{locales}~\cite{Ballarin:2006:MKM}, which allows
us to define named contexts that fix some parameters and assumptions.
For example, the graph theory is developed in the locale \isai{Graph}, which 
fixes the edge labeling function \isai{c}, and defines the set of edges and nodes based on $c$:
\begin{lstlisting}
  locale Graph = fixes c :: "edge => capacity" begin
    definition "E == {(u, v). c (u, v) ~= 0}"
    definition "V == {u. \<exists>v. (u, v) \<in> E \<or> (v, u) \<in> E}"
    [...]
\end{lstlisting}
Moreover, we define basic concepts like (simple, shortest) paths, and provide lemmas to reason about them.

@@

For the Isabelle/HOL proof assistant, the Isar proof language~\cite{Wenzel99} allows to write formal proofs that resemble standard mathematical textbook proofs, and are accessible, to a certain extent, even for those not familiar with Isabelle/HOL. 
We use Isar to present our proof of the Ford-Fulkerson algorithm such that it resembles the informal proof described by Cormen et al.~\cite{CLRS09}.

@@

\subsection{Presentation of Algorithms}
In textbooks, it is common to present algorithms in pseudocode, which captures the essential ideas, but leaves open implementation details. As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}. For example, we define the Ford-Fulkerson algorithm as follows:
\Snippet{ford_fulkerson_algo}

@@

Finally, using the Ford-Fulkerson theorem and the verification condition generator of the Isabelle Refinement Framework, it is straightforward to prove (partial) correctness of
the Ford-Fulkerson algorithm, which is stated in Isabelle/HOL by the following theorem:
\Snippet{ford_fulkerson_correct}

\section{Refinement in Isabelle/HOL}\label{sec:refinement}
After having stated and proved correct an algorithm on the abstract level, the next step is to provide an (efficient) implementation. 
In our case, we first specialize the Ford-Fulkerson algorithm to use shortest augmenting paths, then implement the search for shortest augmenting paths by BFS, and finally use efficient data structures to represent the abstract objects modified by the algorithm.

@@

\section{The Edmonds-Karp Algorithm}\label{sec:edka}
  Specializing the Ford-Fulkerson algorithm to the Edmonds-Karp algorithm is straightforward, as finding a shortest augmenting path is a refinement of finding any augmenting path.

@@

  The push-relabel algorithm maintains a preflow $f$ and a height labeling $l: V \to \mathbb {N}$ of the nodes, 
  with the invariant that for every edge $(u, v)$ of the residual graph we have that $l(u) \le l(v) + 1$. 
  Moreover, the height of the source is fixed to $\left\vert{V}\right\vert$, and the height of the sink is fixed to $0$.
  By generalizing the labeling invariant to paths, we get that for every path $p$ from $u$ to $v$ in the residual graph, we have $l(u) \le l(v) + |p|$.
  This implies that there cannot be an augmenting path, \ie a simple path from $s$ to $t$: the length of any simple path is less than $|V|$, however, a valid labeling 
  implies a length greater than or equal to $|V|$. Upon termination of the algorithm, there will be no active nodes. 
  Thus, the preflow is actually a flow, and as there are no augmenting paths, the min-cut-max-flow theorem implies that the flow is maximal.

@@

The simple structure of the generic algorithm allowed us to conveniently define it as labeled transition system, such that the above informal proof could be formalized
by inductive arguments over the runs of the algorithm.
Moreover, the powerful Isabelle/HOL standard library, in particular the support for summations
over finite sets, was very helpful to keep the formalization effort manageable. 
Nevertheless, the whole complexity proof is roughly 1000 lines of Isabelle text, and took a few weeks to complete.
This includes the time that we spent on an initial termination proof that used simpler arguments, but did not derive an exact complexity bound.

@@

\subsection{FIFO Push-Relabel Algorithm}
While the relabel-to-front algorithm has a theoretical worst case complexity of $O(V^3)$, it seems to be rather inefficient in practice.\footnote{We have not found any implementation based on relabel-to-front, and our own experiments indicate a rather poor performance.}
An alternative implementation is the FIFO push-relabel algorithm, which maintains a queue (FIFO) of the active nodes. 
In each iteration, it dequeues a node and tries to discharge it by only using push operations.
If the node is still active, it is relabeled and enqueued.

The FIFO push-relabel algorithm has several advantages: It's correctness is straightforward to prove, 
as the only required additional invariant is that the queue contains exactly the active nodes. 
Moreover, it performs well in practice, and we have found several reference implementations of it~\cite{Gold_Network_Opt,ICPC-notebook}.

@@

  \subsection{Algorithms}
  The most complex problem for which we specified no algorithm yet is how to find a shortest augmenting path in the Edmonds-Karp algorithm.
  We use breadth first search (BFS) for this. We adapted a BFS formalization from the examples collection of the Isabelle Refinement Framework to
  our needs and added an efficient imperative implementation. The resulting algorithm 
  is independent of flow networks and can be reused for other applications.

@@

  We implement capacities by (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers. Thus, we could chose any executable number representation here.}. Note that an implementation by fixed precision numbers would also be possible,
  but requires additional checks on the network to ensure that no overflows can happen.\footnote{Actually, the unverified reference 
  implementations of the push-relabel algorithms~\cite{Gold_Network_Opt,ICPC-notebook} that we use in our benchmarks (\cf Section~\ref{sec:benchmark}) lack such checks, and silently report erroneous maximum flow values in case of overflow.}
  % TODO: Can we trick the Java reference implementation on integer overflow?
  Nodes are represented by natural numbers less than an upper bound $N$, and residual graphs are implemented by their capacity matrices, which, in turn,
  are realized as arrays of size $N\times N$ with row-major indexing, such that the successors of a node are close together in memory.
  The adjacency map of the network is implemented by an array of lists of nodes. 
  An augmenting path is represented by a list of edges, \ie a list of pairs of nodes. 
  The height frequency map and the maps from nodes to height labels and excess values are implemented by arrays.
  The queue of the FIFO push-relabel algorithm and the list of the relabel-to-front algorithm are implemented by two stacks, allowing for amortized constant time operations.

@@

@@

  For the push-relabel algorithm, we use the C++ implementation from the Stanford ACM-ICPC notebook~\cite{ICPC-notebook}, which served as 
  a template for our verified FIFO push-relabel algorithm. Moreover, we use the Q\_PRF program 
  version $1.8$ from Andrew Goldberg's Network Optimization Library~\cite{Gold_Network_Opt}, which is a highly optimized C implementation of a FIFO push-relabel 
  algorithm with various additional heuristics, originally developed by Cherkassky and Goldberg~\cite{ChGo97}. 
  These were compiled using gcc 5.4 at optimization level 4.
  The data set consists of networks from the DIMACS implementation challenge~\cite{JGC93}: 
  The ``Goldberg bad case networks'' from the ``Washington'' generator, which are tailored to be hard for push-relabel algorithms, 
  the ``fully dense acyclic networks'' of Waissi, and the RMFGEN networks of Goldfarb and Grigoriadis.

@@

  For the push-relabel algorithms, our verified algorithm is, depending on the test case, between 5 and 10 times slower than the unverified C++ implementation.
  The highly optimized C implementation is significantly faster in most cases, only for the ``fully dense acyclic networks'' it is slower than the 
  simple C++ implementation.
  On the ``Goldberg bad case'' networks, which are designed as worst case problems for push-relabel algorithms, 
  even Edmonds-Karp algorithm is competitive to the simple FIFO push-relabel algorithms, while the highly optimized one shows no weaknesses on these networks.
